Implicit yaddayadda

### 2011-09-23 papers/PLDI2012/implicitization.tex
pldi

\subsection{Guided static analysis}
\label{sec:guided_multigraph}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improved over classical upward Kleene iterations with widening.
Consider the following program:
\lstinputlisting{gopan_reps.c}

@@

\section{Guided analysis over the paths}
\label{sec:combined}

Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
graph of the program. One could instead apply the same technique to the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.

Let $P$ be the set of control points in the transition graph, 
$P_W \subseteq P$ the set of widening points
such that removing the points in $P_W$ gives an acyclic graph.
One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.
The combination of these two techniques considers an ascending sequence of
subsets of the paths between two points in $P_R$.

@@

This formula contains Boolean \emph{reachability predicates} $b_i$ for each
control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
that a path 
$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
between two points $p_{i_1}, p_{i_n} \in P_R$ 
can easily be expressed as the
cunjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
$b_i^d$ is $true$ when the path arrives at $p_i$.

The current working subset of paths is stored using a compact representation,
such as binary decision diagrams.

@@

The algorithm for computing an inductive invariant over a subgraph is the Path
Focused algorithm from \citet{Monniaux_Gonnord_SAS11}.

@@

Since we do not work on the entire transition graph but on a subset of it, we
have to cunjunct the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

It is not necessary to add into $P'$ all the new feasible paths. We just have
to add them one after another, until no more path can make the invariant grow,
as proposed by Algorithm \ref{algo:computepaths}. For that, we use another
abstract value associated to a control point $p_j$, noted $X_j^d$, which is
different from $X_j^s$.

@@

We maintain a set of active control points $A$, containing the points in $P_R$
that still need to be processed. We apply the two previously described phases to
the elements of $A$ until $A$ is empty. When $A$ is empty, it means the
invariant computation of the subgraph is finished. As proposed by
\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations, and we
restart the analysis with a new subgraph if needed.
\begin{algorithm}
	\caption{Guided static analysis on implicit multigraph}
	\label{algo:combined}
	\begin{algorithmic}[1] 
	\input{algo/algo-combined}
	\end{algorithmic}
\end{algorithm}

@@

\citet{DBLP:conf/pldi/GulwaniZ10} proposes a technique for computing disjunctive invariants, by
distinguishing all the paths inside a loop. In
this section, we propose to improve this technique by using SMT queries to find
interesting paths, the objective being to avoid an explicit exhaustive
enumeration of an exponential number of paths.

@@

@@

Our improvement consists in a slight modification of the SMT formula we solve.
We introduce in this formula Boolean variables $\{d_k, 1 \leq k \leq m\}$, so
that we can easily find in the model which abstract value of the disjunction we
have to choose. The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.

>>>>

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improved over classical upward Kleene iterations with widening.
Consider the following program:
\lstinputlisting{gopan_reps.c}

@@

\subsection{Path-focused technique}
\label{sec:path_focusing}

\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} proposes to
distinguish the different paths in the program in order to avoid loss of
precision due to merge operations. Since the number of paths may be exponential,
the technique keeps them implicit and computes them when needed using
SMT-solving. Then, the iterations are computed over a reduced multigraph instead
of the classical transition graph.

Let $P$ be the set of control points in the transition graph, 
$P_W \subseteq P$ the set of widening points
such that removing the points in $P_W$ gives an acyclic graph.
One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.

@@

This formula contains Boolean \emph{reachability predicates} $b_i$ for each
control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
that a path 
$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
between two points $p_{i_1}, p_{i_n} \in P_R$ 
can easily be expressed as the
cunjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
a \emph{destination} point, with only incoming transitions, so that the
resulting graph is acyclic and there is no paths going through control
points in $P_R$.

In order to find focus paths, we solve an SMT formula which is satisfiable when
there exist a path starting at a point $p_i \in P_R$ in a state included in the
actual invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
state outside of $X_j$. In this case, we construct this path using the model and
update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
we can apply a widening/narrowing sequence, or even compute the transitive
closure of the loop using abstract acceleration. % TODO citation laure gonnord

\section{Guided analysis over the paths}
\label{sec:guided_multigraph}

Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
graph of the program. One could instead apply the same technique to the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}.

The combination of these two techniques considers an ascending sequence of
subsets of the paths between two points in $P_R$.
The current working subset of paths is stored using a compact representation,
such as binary decision diagrams.

@@

The algorithm for computing an inductive invariant over a subgraph is the Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11}.

@@

Since we do not work on the entire transition graph but on a subset of it, we
cunjunct the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

It is not necessary to add into $P'$ all the new feasible paths. We just have
to add them one after another, until no more path can make the invariant grow,
as proposed by Algorithm \ref{algo:computepaths}. For that, we use another
abstract value associated to a control point $p_j$, noted $X_j^d$, which is
distinct from $X_j^s$, and initialized to $X_j^s$.

@@

We maintain a set of active control points $A$, containing the points in $P_R$
that still need to be processed. We apply the two previously described phases to
the elements of $A$ until $A$ is empty. When $A$ is empty, it means the
invariant computation of the subgraph is finished. As proposed by
\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
These narrowing iterations allow to recover precision lost by
widening, before taking into account new feasible paths.
Thus, our technique combines both the advantages of \emph{Guided Static
Analysis} and \emph{Path-focusing}.
We then restart the analysis with a new subgraph if needed.

\begin{algorithm}
	\caption{Guided static analysis on implicit multigraph}
	\label{algo:combined}
	\begin{algorithmic}[1] 
	\input{algo/algo-combined}
	\end{algorithmic}
\end{algorithm}

@@

\citet{DBLP:conf/pldi/GulwaniZ10} propose a technique for computing disjunctive invariants, by
distinguishing all the paths inside a loop. In
this section, we propose to improve this technique by using SMT queries to find
interesting paths, the objective being to avoid an explicit exhaustive
enumeration of an exponential number of paths.

@@

$m$, $\delta$ and $\sigma$ can be defined heuristically.
For instance, one could defined $\sigma$ so that $\sigma(j,k)$ only depends on the
last transition of the path, or else construct it dynamically during the
analysis.

@@

Our improvement consists in a slight modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_k, 1 \leq k \leq m\}$, so
that we can easily find in the model which abstract value of the disjunction we
have to choose. The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
